Model Order Selection and Eigen Similarity based Framework for Detection and Identification of Network Attacks

### 2016-05-26 docs/papers/2016JNCA/2016JNCA.tex
complexity analysis into paper

\begin{abstract}
Novel schemes for attack detection are crucial to identify adaptive malicious traffic coming from sources that are quickly mobilized by attackers in high throughput communication networks. In this context, signal processing techniques have been applied to attack detection due to their capability to detect anomalies that are previously unknown, i.e. blind detection. This paper proposes a signal processing framework for the detection and identification of network attacks using concepts of model order selection (MOS), eigenvalues and similarity analysis. In order to validate our proposed framework, we consider a network traffic sample that contains malicious activity such as denial of service (DoS) and port probing. We propose to model the network traffic as a superposition of components, namely, user's operations (legitimate traffic), network service operation not related to the user (noise) and the malicious activity. The experiments performed in a real network show that the proposed blind detection approach achieves satisfactory levels of accuracy in terms of timely detection and identification of TCP/UDP ports under attack.
\end{abstract}

@@

Traditionally, cyber defense methods can be effective against ordinary and conventional types of attacks, but may fail against innovative malicious techniques \cite{lakhina2005mining}. In order to be able to detect and avoid novel attacks and their variations, it is necessary to develop or improve techniques to achieve efficiency on resource consumption, processing capacity and response time. Moreover, it is also crucial to obtain high detection accuracy and capacity to detect variations of malicious patterns. Recently, signal processing schemes have being applied to the detection of malicious traffic in computer networks \cite{Lu2009,Huang2009,Zonglin2009,david2011blind,da2012improved,tenorio2013greatest}, showing advances in network traffic analysis.

@@

In the context of anomaly-based schemes, this work proposes an approach for the detection of malicious traffic in computer networks. Inspired by \cite{david2011blind,da2012improved}, this work models the network traffic using a signal processing formulation as a composition of three components: legitimate traffic, malicious traffic and noise, taking into account the incoming and outgoing traffic in certain types of network ports (TCP or UDP). Our proposed technique is based on eigenvalue analysis, model order selection (MOS) and similarity analysis. In contrast to \cite{david2011blind,da2012improved,tenorio2013greatest}, MOS and eigenvalue analysis are applied to detect time frames under attack. In addition, we also evaluate the accuracy of the proposed framework. Additionally, this proposed approach has its accuracy evaluation based on eigen similarity analysis for extracting detailed information about accurate time and network ports under attack.

The performed experiments show that synflood, fraggle and port scan attacks can be detected accurately and with great detail in an automatic and blind fashion, applying signal processing concepts for traffic modeling and through approaches based on MOS and eigen similarity analysis. The main contributions of our proposed framework are the capability to blindly detect time frames under network attack via MOS and eigen analysis, and the detailed identification of the network attack via eigen similarity analysis.

This paper is organized as follows. In Section \ref{sec:relatedworks}, related works are discussed. Section \ref{sec:datamodel} presents the data model. Section \ref{sec:prop_getv} describes our proposed framework for blind and automatic detection of malicious traffic. Section \ref{sec:experimentalresults} discusses the experimental validation and presents the experimental results. Section \ref{sec:conclusionandfutureworks} draws the conclusions and the suggestions for future work. The \ref{sec:mos} presents mathematical concepts of examples of state-of-the-art MOS schemes.

@@

Signal processing techniques have been successfully applied to network anomaly detection \cite{Lu2009}. Lu and Ghorbani \cite{Lu2009} proposed a network anomaly detection model based on network flow, wavelet approximation, and system identification theory. However, their work did not address problems without significant outliers, such as port scan attacks. Zonglin \emph{et al} \cite{Zonglin2009} proposed a method to detect traffic anomaly with correlation analysis, where the correlation between traffic signals and the predicted traffic signals are used to reveal anomalies. Zonglin \emph{et al} \cite{Zonglin2009} evaluated the correlation analysis for anomaly detection, but the work was not applied to probing and denial of service (DoS) attack detection, simultaneously.

@@

Lee \emph{et al} \cite{Lee2013} proposed OverSampling PCA (osPCA), which allows one to determine the anomaly of the target instance according to the variation of the resulting dominant eigenvector obtained by similarity analysis and over sampling. In contrast to Lee \emph{et al}, our framework applies MOS for detection of time frames under attack and similarity analysis to extract details for detection of time and ports under attack. Additionally, Lee \emph{et al} only evaluated their proposed scheme for covariance analysis, while we adopted an analysis based on sample covariance of zero mean variables and sample covariance of zero mean and unitary standard deviation variables, for DoS and probing attacks, respectively.

Our proposed framework does not require either a significant amount of logs to detect attacks, nor prior data collection, in order to make comparisons and evaluate the existence of malicious traffic. Our proposed solution is automatic and blind for detection of time frames under probing and DoS attacks through MOS and eigen analysis. Moreover, we apply eigen similarity analysis to identify details of time and ports under network attacks.

@@

In this paper, scalars are denoted by italic letters (\emph{a, b, A, B, $α$, $β$}), vectors by lowercase bold letters (\textbf{a, b}), matrices by uppercase bold letters (\textbf{A, B}), and $a_i,_j$ denotes the (\emph{i, j}) elements of the matrix \textbf{A}. The superscripts \textsuperscript{T} and \textsuperscript{-1} are used for matrix transposition and matrix inversion, respectively. We define the operator $\rm diag(\cdot)$ that returns the vector of the main diagonal of a given matrix, the operator $\rightarrow$, which denotes the deletion of a given element from a set and the operator $\#$, that returns the rank of a matriz.

This section presents details of the simulated scenario along with a description of the dataset model as a signal superposition of legitimate traffic, noise and malicious traffic. Subsection \ref{sec:DataCollection} describes the environment and scenario adopted in order to reproduce DoS and probing attacks. Subsection \ref{sec:ModelingData} presents how network traffic can be modelled as signal superposition, and Subsection \ref{sec:AnalyzedScenario} details the traffic of synflood, fraggle and port scan attacks.

@@

The environment of the analyzed scenario is composed of two computers and one router with access to Internet and to an internal network, where the simulation of legitimate traffic, noise, DoS and port scan attacks are performed. During the traffic generation, one computer assumes the role of attacker, while the other is the victim, according to scenario represented by Figure \ref{fig:fig1}.

@@

In the proposed framework, the goal is to detect the anomalies only taking into account the traffic profile, i.e. specific information such as origin IP or day and time of the attack are not considered. Therefore, from the entire log information, we just consider the timestamp (for sequencing), port type and port number.

@@

\begin{equation}\label{eq:eq01}
\boldsymbol{X}^{(q)} = \boldsymbol{U}^{(q)} + \boldsymbol{N}^{(q)} + \boldsymbol{A}^{(q)},
\end{equation}
where $q$ represents the $q$-th time frame, which is a time grouping of network traffic. The matrix $\boldsymbol{X}^{(q)} \in \mathbb{R}^{M \times N}$ consists of \emph{M} rows and \emph{N} columns. Each row represents a communication port (TCP port or UDP port), and each column represents the time, whose unit can be minutes, for instance. Each element $x_{m,n}^{(q)}$ stands for the number of times that the port $m$ appears at the $n$-th minute, at the $q$-th time frame.

@@

The traffic that is not associated with user's operations and with malicious traffic is modelled as noise $\boldsymbol{N}^{(q)}$. The automatic acquisition service of logical IP network address (DHCP) is an example of noise. Independently of any user operation, the machine receives an IP address, since it is configured to perform a DHCP address request. Figure \ref{fig:fig4} depicts an example of noise in a network traffic, represented by traffic to ports 67 and 68.

@@

With respect to the fraggle atack, large packets with UDP echo segments are sent to the broadcast address of a network. Every packet is modified to have the source address of the victim, in order to implement the source address spoofing technique. Therefore, each host receives a huge amount of requests UDP echo and all of them replies to the IP address of the victim, causing a packet flooding aiming a DoS. This attack can affect the entire network, since all hosts receive several requests UDP echo and respond with the ICMP protocol, therefore each host acts as an amplifier of the attack. This last part of the fraggle attack is not taken into account in this work, because the victim receives ICMP (network layer) packets originated from the hosts that are attacked with flooding packet UDP echo. Figure \ref{fig:fig6} depicts an example of the fraggle attack in a real computer network.

@@

According to DoS and port scan attacks' behavior, DoS attacks and port scan attacks can be characterized as covariance aware attack \citep{jin2004covariance} and correlation aware attack \citep{lakhina2005mining}, respectively. These characteristics are substantiated by the results obtained through our analysis based on sample covariation of zero mean variables and on covariance of zero mean and unitary standard deviation variables, described in Section \ref{sec:experimentalresults}, which shows that the main components of DoS attacks are dominated by the variables with more variance and that the traffic associated with port scan attack does not generate many logs, however, it presents high covariance of zero mean and unitary standard deviation variables.

@@

The eigenvalues should be sorted in descending order, i.e. $\lambda_{1}^{(q)} > \lambda_{2}^{(q)} > \lambda_{3}^{(q)} > ... > \lambda_{m}^{(q)}$. Therefore, the largest eigenvalue of the $q$-th time frame evaluated for the attack detect is given by $\lambda_{1}^{(q)}$.

@@

Traditionally the MOS schemes are applied for the eigenvalues of the vector $\boldsymbol{e}^{(q)}$. However, our goal here is to detect the variations of the eigenvalues for different values of $q$. Therefore, instead of using a certain $q$, our approach applies MOS schemes for a vector of the largest eigenvalues of each $q$-th time frame, in order to identify variations and estimate the model order $\hat{d}$, which is the estimated number of time frames under attack. Therefore, $\boldsymbol{e}_{\rm max}$ is sorted in descending order, producing $\sim\boldsymbol{e}_{\rm max}$, that is used as input parameter for MOS schemes, according to $\hat{d} = \rm{MOS}(\sim\boldsymbol{e}_{\rm max})$. Note that some MOS schemes may also require the number of minutes that compose a time frame, as $\hat{d} = \rm{MOS}(\boldsymbol{e}_{\rm max},\emph{Q})$. For more information about MOS, we refer to \ref{sec:mos}.

@@

After applying the MOS schemes to the vector $\sim\boldsymbol{e}_{\rm max}$, we obtain the an estimate of the $\#\boldsymbol{A}$. For instance, in the case of fraggle, synflood and portscan, if $\hat{d} = 1$, , then $\#\boldsymbol{A} = 1$, which means that during the during the $Q$ time frames one attack is present. However, if $\hat{d} = 0$, then $\#\boldsymbol{A} = 0$, and this means that none of these attacks are present. Note that $\hat{d}$ can be greater than 1, indicating the presence of more than one attack.

In Subsection \ref{sec:prop_MOSSchemes}, we obtained only if $\hat{d} = 1$ or $\hat{d} = 0$. However, if $\hat{d} = 1$, the MOS schemes do not provide any information about the $q$-th time frame under attack. The identification of the $q$-th time frame under attack can be done through a eigenvalues analysis.

@@

This subsection describes the proposed eigen similarity analysis for detailed attack identification, in complement to the attack estimation done through MOS schemes and eigenvalue analysis. In Subsection \ref{sec:prop_TimeSimilarityAnalysis} we present the eigen similarity analysis for identification of time under attack. Next, in Subsection \ref{sec:prop_PortSimilarityAnalysis}, we show how to apply the eigen similarity analysis in order to identify network ports under attack.

@@

\begin{equation}\label{eq:eq11}
s_n = \abs{\frac{\boldsymbol{v}^{(q)} \cdot \boldsymbol{v}_{(n)}}{\norm{\boldsymbol{v}^{(q)}}\norm{\boldsymbol{v}_{(n)}}}},
\end{equation}
where $s_n$ denotes the absolute similarity degree of the $n$-th minute, $\boldsymbol{v}^{(q)}$ is the most significant eigenvectors of a selected set of minutes without network attack, and $\boldsymbol{v}_{(n)}$ is the most significant eigenvectors obtained after append the target $n$-th minute of traffic to be performed the DoS and port scan attack identification.

@@

The Figure \ref{fig:fig8} ilustrates the network traffic selection for the incremental approach of eigen similarity analysis, where the $\boldsymbol{X}^{(1)}$ is chosen as reference for similarity analysis of the $m$-th minutes of the time frame $q=3$, where one network attack was previously detected.

@@

The eigen similarity analysis starts at $\boldsymbol{x}^{(\hat{3})}_{(41)}$ and is incrementally performed until $\boldsymbol{x}^{(\hat{3})}_{(60)}$, in order to calculate the $s_n$. We assume that $s_n < l$ means an attack identification, according the anomaly on similiarity of $s_n$ to a defined limiar $l$. Therefore, after obtaining the most significant eigenvector $\boldsymbol{v}^{(q)}$ and the target eigenvectors $\boldsymbol{v}_{(n)}$ for eigen similarity analyis, the $s_n$ is calculated according to (\ref{eq:eq11}).

@@

\begin{equation}\label{eq:eq15}
  \left\{
  \begin{array}{@{}ll@{}}
    x_{(m,n)} = x^{(\hat{q})}_{(m,\hat{n})} \\
    \\
    s_{m,\hat{n}} = \abs{\frac{\boldsymbol{v}^{(q)} \cdot \boldsymbol{v}_{(m,\hat{n})}}{\norm{\boldsymbol{v}^{(q)}}\norm{\boldsymbol{v}_{(m,\hat{n})}}}},
  \end{array}\right.
\end{equation}
where $x^{(\hat{q})}_{(m,\hat{n})}$ denotes the $m$-th port of the selected $n$-th minute identified as under attack and $x_{(m,n)}$ denotes the $m$-th port of the last $n$-th minute of $\boldsymbol{X}_n$, which is used to calculate the $\boldsymbol{v}_{(m,\hat{n})}$ most significant eigenvectors that contains the traffic of the $m$-th port of the $\hat{n}$-th minute identified as under attack.

@@

This section presents the performed experiments and the acquired results. First, in Section \ref{sec:AnalyzedScenario}, the scenario adopted in the experiments is summarized. Then, Section \ref{sec:largesteigenvaluesanalysis} shows the results of the largest eigenvalue analysis by time frames. In Section \ref{sec:MOSSchemesEvaluation} are described the results of the evaluated MOS schemes for attack detection. Section \ref{sec:EigenvalueAnalysis} presents the results of the eigenvalue analysis for identification of time frames under attack, and Section \ref{sec:EigenSimilarityAnalysis} shows the results of similarity analysis for detailed DoS and port scan identification.

@@

For the evaluation of MOS Schemes acuracy for DoS and port scan detection, our framework defines that it is necessary to obtain the largest eigenvalue of each time frame, through eigen decomposition from a covariance of zero mean variables or covariance matrix of zero mean and unitary standard deviation variables, calculated from the evaluated traffic, as described in Section \ref{sec:prop_getv}. Through eigenvalue analysis of traffic with DoS or port scan attacks, it is possible to visualize a significant difference between the largest eigenvalues and the remain eigenvalues, which can indicate a relationship between an outlier and time frames under attack.

@@

These results highlight that all $q$-th time frames, where a network attack was simulated, present high significant variance between the largest eigenvalue and the remaining eigenvalues, obtained from sample covariance matrix, for DoS detection, or from covariance matrix of zero mean and unitary standard deviation variables, for port scan detection. Therefore, we propose to apply the vector of the largest eigenvalues to MOS schemes in order to evaluate their accuracy for identification of time frames under attack, motivated by the fact that it is relevant to apply MOS schemes to automate the attack detection process, taking into account the characteristics of the evaluated eigenvalues.

@@

Although EDC and EFT presented the same accuracy on our evaluation, the EDC scheme requires less processing time than EFT, which is an important criteria to select EDC as the MOS scheme for DoS and port scan detection on our remain experiments.

@@

The Table \ref{tab:tab5} presents the results of the evaluation of three approaches for similarity analysis of eigenvectors for port scan detection.

@@

The Table \ref{tab:tab5} shows the evaluation of the time frame $q=3$, when the port scan attack was simulated, considering the incremental individualized, incremental and individual approaches for eigen similarity analysis. According to the presented results, it is possible to observe the high similarity between network traffic without attack, which was larger than 0.9610 for all evaluated cases, and emphasize the expressive low similarity when evaluated the traffic with the simulated port scan attack ($n=15$), which was lower than 0.0276 for all evaluated approaches.

Comparing the approaches for similarity analysis, it is possible to observe that all approaches highlight the low similarity when evaluated the traffic under attack. However, the incremental approach figured out low similarity for times without attack, where $n=16, 17, 18, 19, 20$, what indicates that the incremental approach can produce false posive results. This behaviour occurs because the incremental approaches appends all selected traffic into the reference traffic for comparison against the original reference traffic, what makes more evident the first lack of similarity but reduces the changing detection capability after an attack detection.

The Table \ref{tab:tab6} presents the results of the evaluation of the similarity analysis of eigenvectors for synflood detection. It shows the evaluation of the time frame $q=4$, when the synflood attack is simulated, considering the incremental individualized, incremental and individual approaches for eigen similarity analysis. According to the  results, it is possible to observe the high similarity between network traffic without attack, which is larger than 0.9907 for all evaluated cases, and emphasize the expressive low similarity when evaluated the traffic with synflood attack (between $n=11$ and $n=20$), which is lower than 0.1244 for all evaluated approaches.

@@

Comparing the incremental individualized and the individual approaches for eigen similarity analysis, it is possible to observe that the incremental individualized approach obtain lowest values for almost all cases, except for the time $n=14$, where incremental individualized approach identified a larger similarity than the individual approach. The incremental individualized appends information about each evaluated traffic, therefore it incorporates traffic behaviours that can reduce the outlier capability detection, as ocurred for the time $n=14$.

The Table \ref{tab:tab7} presents the results of the eigen similarity analysis evaluation for fraggle detection.

@@

For fraggle attack detection, the lack of similarity between legitimate and malicious traffic was more evident than for the evaluation of synflood and port scan detecion. This becaviour can be explained by the number of packets generated through the fraggle attack simulation, that was significative larger than the number of packets generated during the synflood simulation. Considering the three approaches, the largest value for times under attack was 0.0083, while the shortest value for times without attacks was 0.9993.

@@

The Table \ref{tab:tab8} presents the results of the evaluation of eigen similarity analysis for detection of ports under port scan attack, showing only the time frame $q=3$ and minute $n=15$, due to the simulated port scan attack occurred only at this time, although the remain time frame has been completely evaluated and presented high similarity to the reference of traffic without network attack.

>>>>

\begin{abstract}
Novel schemes for attack detection are crucial to identify adaptive malicious traffic coming from sources that are quickly mobilized by attackers in high throughput communication networks. In this context, signal processing techniques have been applied to attack detection due to their capability to detect anomalies that are previously unknown, i.e. blind detection. This paper proposes a signal processing framework for the detection and identification of network attacks using concepts of model order selection (MOS), eigenvalues and similarity analysis. In order to validate the proposed framework, we consider a network traffic sample that contains malicious activity such as denial of service (DoS) and port probing. We propose to model the network traffic as a superposition of components, namely, user's operations (legitimate traffic), network service operation not related to the user (noise) and the malicious activity. The experiments performed in a real network show that the proposed blind detection approach achieves satisfactory levels of accuracy in terms of timely detection and identification of TCP/UDP ports under attack.
\end{abstract}

@@

Traditionally, cyber defense methods can be effective against ordinary and conventional types of attacks, yet may fail against innovative malicious techniques \cite{lakhina2005mining}. In order to be able to detect and avoid novel attacks and their variations, it is necessary to develop or improve techniques to achieve efficiency on resource consumption, processing capacity and response time. Moreover, it is crucial to obtain high detection accuracy and capacity to detect variations of malicious patterns. Recently, signal processing schemes have being applied to the detection of malicious traffic in computer networks \cite{Lu2009,Huang2009,Zonglin2009,david2011blind,da2012improved,tenorio2013greatest}, showing advances in network traffic analysis.

@@

In the context of anomaly-based schemes, this work proposes an approach for the detection of malicious traffic in computer networks. Inspired by \cite{david2011blind,da2012improved}, this work models the network traffic using a signal processing formulation as a composition of three components: legitimate traffic, malicious traffic and noise, taking into account the incoming and outgoing traffic in certain types of network ports (TCP or UDP). To the best of our knowledge there is no similar model in the literature. The proposed technique is based on eigenvalue analysis, model order selection (MOS) and similarity analysis. In contrast to \cite{david2011blind,da2012improved,tenorio2013greatest}, MOS and eigenvalue analysis are applied to detect time frames under attack. In addition, we also evaluate the accuracy of the proposed framework. Additionally, this proposed approach has its accuracy evaluation based on eigen similarity analysis for extracting detailed information about accurate time and network ports under attack.

The performed experiments show that synflood, fraggle and port scan attacks can be detected accurately and with great detail in an automatic and blind fashion, applying signal processing concepts for traffic modeling and through approaches based on MOS and eigen similarity analysis. The main contributions of the proposed framework are the capability to blindly detect time frames under network attack via MOS and eigen analysis, and the detailed identification of the network attack via eigen similarity analysis.

This paper is organized as follows. In Section \ref{sec:relatedworks}, related works are discussed. Section \ref{sec:datamodel} presents the data model. Section \ref{sec:prop_getv} describes the proposed framework for blind and automatic detection of malicious traffic. Section \ref{sec:experimentalresults} discusses the experimental validation and presents the experimental results. Section \ref{sec:conclusionandfutureworks} draws the conclusions and the suggestions for future work. The \ref{sec:mos} presents mathematical concepts of examples of state-of-the-art MOS schemes.

@@

Signal processing techniques have been successfully applied to network anomaly detection \cite{Lu2009}. Lu and Ghorbani \cite{Lu2009} proposed a network anomaly detection model based on network flow, wavelet approximation, and system identification theory. However, their work did not address problems without significant outliers, such as port scan attacks. Zonglin \emph{et al} \cite{Zonglin2009} proposed a method to detect traffic anomaly with correlation analysis, where the correlation between traffic signals and the predicted traffic signals are used to reveal anomalies. Zonglin \emph{et al.} \cite{Zonglin2009} evaluated the correlation analysis for anomaly detection, but the work was not applied to probing and denial of service (DoS) attack detection, simultaneously.

@@

Lee \emph{et al.} \cite{Lee2013} proposed OverSampling PCA (osPCA), which allows one to determine the anomaly of the target instance according to the variation of the resulting dominant eigenvector obtained by similarity analysis and over sampling. In contrast to Lee \emph{et al.}, the framework applies MOS for detection of time frames under attack and similarity analysis to extract details for detection of time and ports under attack. Additionally, Lee \emph{et al.} only evaluated their proposed scheme for covariance analysis, while we adopted an analysis based on sample covariance of zero mean variables and sample covariance of zero mean and unitary standard deviation variables, for DoS and probing attacks, respectively.

The proposed framework does not require either a significant amount of logs to detect attacks, nor prior data collection, in order to make comparisons and evaluate the existence of malicious traffic. The proposed solution is automatic and blind for detection of time frames under probing and DoS attacks through MOS and eigen analysis. Moreover, we apply eigen similarity analysis to identify details of time and ports under network attacks.

@@

In this paper, scalars are denoted by italic letters (\emph{a, b, A, B, $α$, $β$}), vectors by lowercase bold letters (\textbf{a, b}), matrices by uppercase bold letters (\textbf{A, B}), and $a_i,_j$ denotes the (\emph{i, j}) elements of the matrix \textbf{A}. The superscripts \textsuperscript{T} and \textsuperscript{-1} are used for matrix transposition and matrix inversion, respectively. We define the operator $\rm diag(\cdot)$ that returns the vector of the main diagonal of a given matrix, the operator $\rightarrow$, which denotes the deletion of a given element from a set and the operator $\#$, that returns the rank of a matriz, and the operator $\sim$ that sorts the elements of a vector in ascending order.

This section presents details of the simulated scenario along with a description of the dataset model as a signal superposition of legitimate traffic, noise and malicious traffic. Subsection \ref{sec:DataCollection} describes the environment and scenario adopted in order to reproduce DoS and probing attacks. Subsection \ref{sec:ModelingData} presents how network traffic can be modeled as signal superposition, and Subsection \ref{sec:AnalyzedScenario} details the traffic of synflood, fraggle and port scan attacks.

@@

The environment of the analyzed scenario is composed of two computers and one router with access to the Internet and to an internal network, where the simulation of legitimate traffic, noise, DoS and port scan attacks are performed. During the traffic generation, one computer assumes the role of the attacker, while the other is the victim, according to scenario represented by Figure \ref{fig:fig1}.

@@

In the proposed framework, the goal is to detect the anomalies only taking into account the traffic profile, i.e., specific information such as origin IP or day and time of the attack are not considered. Therefore, from the entire log information, we just consider the timestamp (for sequencing), port type and port number.

@@

\begin{equation}\label{eq:eq01}
\boldsymbol{X}^{(q)} = \boldsymbol{U}^{(q)} + \boldsymbol{N}^{(q)} + \boldsymbol{A}^{(q)},
\end{equation}
where $q$ represents the $q$-th time frame, which is a time grouping of network traffic. The matrix $\boldsymbol{X}^{(q)} \in \mathbb{R}^{M \times N}$ consists of \emph{M} rows and \emph{N} columns. Each row represents a communication port (TCP port or UDP port), and each column represents time bins having a approtiate size, such as one minute. Each element $x_{m,n}^{(q)}$ stands for the number of times that the port $m$ appears at the $n$-th minute, at the $q$-th time frame.

@@

The traffic that is not associated with user's operations and with malicious traffic is modeled as noise $\boldsymbol{N}^{(q)}$. The automatic acquisition service of logical IP network address (DHCP) is an example of noise. Independently of any user operation, the machine receives an IP address, since it is configured to perform a DHCP address request. Figure \ref{fig:fig4} depicts an example of noise in a network traffic, represented by traffic to ports 67 and 68.

@@

With respect to the fraggle attack, large packets with UDP echo segments are sent to the broadcast address of a network. Every packet is modified to have the source address of the victim, in order to implement the source address spoofing technique. Therefore, each host receives a huge amount of requests UDP echo and all of them replies to the IP address of the victim, causing a packet flooding aiming a DoS. This attack can affect the entire network, since all hosts receive several requests UDP echo and respond with the ICMP protocol, therefore each host acts as an amplifier of the attack. This last part of the fraggle attack is not taken into account in this work, because the victim receives ICMP (network layer) packets originated from the hosts that are attacked with flooding packet UDP echo. Figure \ref{fig:fig6} depicts an example of the fraggle attack in a real computer network.

@@

According to DoS and port scan attacks' behavior, DoS attacks and port scan attacks can be characterized as covariance aware attack \citep{jin2004covariance} and correlation aware attack \citep{lakhina2005mining}, respectively. These characteristics are substantiated by the results obtained through the analysis based on sample covariation of zero mean variables and on covariance of zero mean and unitary standard deviation variables, described in Section \ref{sec:experimentalresults}, which shows that the main components of DoS attacks are dominated by the variables with more variance and that the traffic associated with port scan attack does not generate many logs, however, it presents high covariance of zero mean and unitary standard deviation variables.

@@

The eigenvalues should be sorted in descending order, i.e., $\lambda_{1}^{(q)} > \lambda_{2}^{(q)} > \lambda_{3}^{(q)} > ... > \lambda_{m}^{(q)}$. Therefore, the largest eigenvalue of the $q$-th time frame evaluated for the attack detect is given by $\lambda_{1}^{(q)}$.

@@

Traditionally the MOS schemes are applied for the eigenvalues of the vector $\boldsymbol{e}^{(q)}$. However, the goal here is to detect the variations of the eigenvalues for different values of $q$. Therefore, instead of using a certain $q$, the proposed approach applies MOS schemes for a vector of the largest eigenvalues of each $q$-th time frame, in order to identify variations and estimate the model order $\hat{d}$, which is the estimated number of time frames under attack. Therefore, $\boldsymbol{e}_{\rm max}$ is sorted in descending order, producing $\sim\boldsymbol{e}_{\rm max}$, that is used as input parameter for MOS schemes, according to $\hat{d} = \rm{MOS}(\sim\boldsymbol{e}_{\rm max})$. Note that some MOS schemes may also require the number of minutes that compose a time frame, as $\hat{d} = \rm{MOS}(\boldsymbol{e}_{\rm max},\emph{Q})$. For more information about MOS, we refer to \ref{sec:mos}.

@@

After applying the MOS schemes to the vector $\sim\boldsymbol{e}_{\rm max}$, we obtain the estimate of the $\#\boldsymbol{A}$. For instance, in the case of fraggle, synflood and portscan, if $\hat{d} = 1$, , then $\#\boldsymbol{A} = 1$, which means that during the during the $Q$ time frames one attack is present. However, if $\hat{d} = 0$, then $\#\boldsymbol{A} = 0$, and this means that none of these attacks are present. Note that $\hat{d}$ can be greater than 1, indicating the presence of more than one attack.

In Subsection \ref{sec:prop_MOSSchemes}, we obtained only if $\hat{d} = 1$ or $\hat{d} = 0$. However, if $\hat{d} = 1$, the MOS schemes do not provide any information about the $q$-th time frame under attack. The identification of the $q$-th time frame under attack can be carried out through a eigenvalues analysis.

@@

This subsection describes the proposed eigen similarity analysis for detailed attack identification, in complement to the attack estimation carried out through MOS schemes and eigenvalue analysis. In Subsection \ref{sec:prop_TimeSimilarityAnalysis} we present the eigen similarity analysis for identification of time under attack. Next, in Subsection \ref{sec:prop_PortSimilarityAnalysis}, we show how to apply the eigen similarity analysis in order to identify network ports under attack.

@@

\begin{equation}\label{eq:eq11}
s_n = \frac{\abs{\boldsymbol{v}^{(q)} \cdot \boldsymbol{v}_{(n)}}}{\norm{\boldsymbol{v}^{(q)}}\norm{\boldsymbol{v}_{(n)}}},
\end{equation}
where $s_n$ denotes the absolute similarity degree of the $n$-th minute, $\boldsymbol{v}^{(q)}$ is the most significant eigenvectors of a selected set of minutes without network attack, and $\boldsymbol{v}_{(n)}$ is the most significant eigenvectors obtained after append the target $n$-th minute of traffic to be performed the DoS and port scan attack identification.

@@

Figure \ref{fig:fig8} ilustrates the network traffic selection for the incremental approach of eigen similarity analysis, where the $\boldsymbol{X}^{(1)}$ is chosen as reference for similarity analysis of the $m$-th minutes of the time frame $q=3$, where one network attack was previously detected.

@@

The eigen similarity analysis starts at $\boldsymbol{x}^{(3)}_{(41)}$ and is incrementally performed until $\boldsymbol{x}^{(\hat{3})}_{(60)}$, in order to calculate the $s_n$. We assume that $s_n < l$ means an attack identification, according the anomaly on similiarity of $s_n$ to a defined limiar $l$. Therefore, after obtaining the most significant eigenvector $\boldsymbol{v}^{(q)}$ and the target eigenvectors $\boldsymbol{v}_{(n)}$ for eigen similarity analyis, the $s_n$ is calculated according to~(\ref{eq:eq11}).

@@

\begin{equation}\label{eq:eq15}
  \left\{
  \begin{array}{@{}ll@{}}
    x_{(m,n)} = x^{(\hat{q})}_{(m,\hat{n})} \\
    \\
    s_{m,\hat{n}} = \frac{\abs{\boldsymbol{v}^{(q)} \cdot \boldsymbol{v}_{(m,\hat{n})}}}{\norm{\boldsymbol{v}^{(q)}}\norm{\boldsymbol{v}_{(m,\hat{n})}}},
  \end{array}\right.
\end{equation}
where $x^{(\hat{q})}_{(m,\hat{n})}$ denotes the $m$-th port of the selected $n$-th minute identified as under attack and $x_{(m,n)}$ denotes the $m$-th port of the last $n$-th minute of $\boldsymbol{X}_n$, which is used to calculate the $\boldsymbol{v}_{(m,\hat{n})}$ most significant eigenvectors that contains the traffic of the $m$-th port of the $\hat{n}$-th minute identified as under attack.

@@

This section presents the performed experiments and the acquired results. First, in Section \ref{sec:AnalyzedScenario}, the scenario adopted in the experiments is summarized. Then, Section \ref{sec:largesteigenvaluesanalysis} shows the results of the largest eigenvalue analysis by time frames. In Section \ref{sec:MOSSchemesEvaluation} are described the results of the evaluated MOS schemes for attack detection. Section \ref{sec:EigenvalueAnalysis} presents the results of the eigenvalue analysis for identification of time frames under attack, Section \ref{sec:EigenSimilarityAnalysis} shows the results of similarity analysis for detailed DoS and port scan identification, and Section \ref{sec:Complexity} discuss the computational complexity of the proposed framework.

@@

For the evaluation of MOS Schemes acuracy for DoS and port scan detection, the framework defines that it is necessary to obtain the largest eigenvalue of each time frame, through eigen decomposition from a covariance of zero mean variables or covariance matrix of zero mean and unitary standard deviation variables, calculated from the evaluated traffic, as described in Section \ref{sec:prop_getv}. Through eigenvalue analysis of traffic with DoS or port scan attacks, it is possible to visualize a significant difference between the largest eigenvalues and the remain eigenvalues, which can indicate a relationship between an outlier and time frames under attack.

@@

These results highlight that all $q$-th time frames where a network attack was simulated, present high significant variance between the largest eigenvalue and the remaining eigenvalues, obtained from sample covariance matrix, for DoS detection, or from covariance matrix of zero mean and unitary standard deviation variables, for port scan detection. Therefore, we propose to apply the vector of the largest eigenvalues to MOS schemes in order to evaluate their accuracy for identification of time frames under attack, motivated by the fact that it is relevant to apply MOS schemes to automate the attack detection process, taking into account the characteristics of the evaluated eigenvalues.

@@

Although EDC and EFT presented the same accuracy on the evaluation, the EDC scheme requires less processing time than EFT, which is an important criteria to select EDC as the MOS scheme for DoS and port scan detection on the remain experiments.

@@

Table \ref{tab:tab5} presents the results of the evaluation of three approaches for similarity analysis of eigenvectors for port scan detection.

@@

Table \ref{tab:tab5} shows the evaluation of the time frame $q=3$, when the port scan attack was simulated, considering the incremental individualized, incremental and individual approaches for eigen similarity analysis. According to the presented results, it is possible to observe the high similarity between network traffic without attack, which was larger than 0.9610 for all evaluated cases, and emphasize the expressive low similarity when evaluated the traffic with the simulated port scan attack ($n=15$), which was lower than 0.0276 for all evaluated approaches.

Comparing the approaches for similarity analysis, it is possible to observe that all approaches highlight the low similarity when evaluated the traffic under attack. However, the incremental approach figured out low similarity for times without attack, where $n=16, 17, 18, 19, 20$, what indicates that the incremental approach can produce false posive results. This behavior occurs because the incremental approaches appends all selected traffic into the reference traffic for comparison against the original reference traffic, what makes more evident the first lack of similarity but reduces the changing detection capability after an attack detection.

Table \ref{tab:tab6} presents the results of the evaluation of the similarity analysis of eigenvectors for synflood detection. It shows the evaluation of the time frame $q=4$, when the synflood attack is simulated, considering the incremental individualized, incremental and individual approaches for eigen similarity analysis. According to the  results, it is possible to observe the high similarity between network traffic without attack, which is larger than 0.9907 for all evaluated cases, and emphasize the expressive low similarity when evaluated the traffic with synflood attack (between $n=11$ and $n=20$), which is lower than 0.1244 for all evaluated approaches.

@@

Comparing the incremental individualized and the individual approaches for eigen similarity analysis, it is possible to observe that the incremental individualized approach obtain lowest values for almost all cases, except for the time $n=14$, where incremental individualized approach identified a larger similarity than the individual approach. The incremental individualized appends information about each evaluated traffic, therefore it incorporates traffic behaviors that can reduce the outlier capability detection, as ocurred for the time $n=14$.

Table \ref{tab:tab7} presents the results of the eigen similarity analysis evaluation for fraggle detection.

@@

For fraggle attack detection, the lack of similarity between legitimate and malicious traffic was more evident than for the evaluation of synflood and port scan detecion. This behavior can be explained by the number of packets generated through the fraggle attack simulation, that was significative larger than the number of packets generated during the synflood simulation. Considering the three approaches, the largest value for times under attack was 0.0083, while the shortest value for times without attacks was 0.9993.

@@

Table \ref{tab:tab8} presents the results of the evaluation of eigen similarity analysis for detection of ports under port scan attack, showing only the time frame $q=3$ and minute $n=15$, due to the simulated port scan attack occurred only at this time, although the remain time frame has been completely evaluated and presented high similarity to the reference of traffic without network attack.
